### 2023-06-22 tex/syntax.tex
[ add ] graph example

>>>>

\subsection{An encoding of graphs}

As a non-logical example of a syntax that can be encoded using linear syntax
descriptions, I consider a language of directed acyclic hypergraphs, which can
be used as string diagrams for symmetric monoidal categories.
I want to represent hypergraphs like the following, made up of operators
$\{
\tikz{\node (n) [circle,draw,minimum size=2mm,inner sep=0] {};%
  \draw[->] (n) to (0,-3mm)},
\tikz{\node (n) [circle,draw,minimum size=2mm,inner sep=0] {};%
  \draw[->] (2.12mm,2.12mm) to (n); \draw[->] (-2.12mm,2.12mm) to (n);%
  \draw[->] (n) to (0,-3mm)},
\tikz{\node (n) [circle,fill,minimum size=2mm,inner sep=0] {};%
  \draw[->] (0,3mm) to (n)},
\tikz{\node (n) [circle,fill,minimum size=2mm,inner sep=0] {};%
  \draw[->] (n) to (2.12mm,-2.12mm); \draw[->] (n) to (-2.12mm,-2.12mm);%
  \draw[->] (0,3mm) to (n)}
\}$
and wires between them with no fan-in or fan-out.
I have given the wires names (a, b, w, x, y, and z), which I use in a textual
representation of the graph on the right.
The textual representation is in A-normal form, where the two nullary operators
have names beginning with \texttt{eta}, the two binary operators have names
beginning with \texttt{mu}, and names having the suffix \texttt{white} or
\texttt{black} based on the corresponding operator's colour.

\noindent
\begin{minipage}[t]{0.5\textwidth}
  \begin{tikzpicture}[baseline]
    \node (i0) {};
    \node[right=2cm of i0] (i1) {};
    \node[below=1cm of i0] (mub) [circle,fill] {};
    \node[below right=1cm of mub] (muw) [circle,draw] {};
    \node[below left=1cm of muw] (etaw) [circle,draw] {};
    \node[below right=1cm of etaw] (etab) [circle,fill] {};
    \node[below left=1cm of etab] (o0) {};
    \node[below right=1cm of etab] (o1) {};
    \draw[->] (i0) to node[auto] {a} (mub);
    \draw[->] (i1) to[out=-90,in=45] node[auto] {b} (muw);
    \draw[->] (mub) to node[auto] {x} (muw);
    \draw[->] (muw) to node[above right] {y} (etab);
    \draw[->] (etaw) to node[auto] {z} (o0);
    \draw[->] (mub) to[out=-135,in=90] node[left,near start] {w} (o1);
  \end{tikzpicture}
\end{minipage}
\begin{minipage}[t]{0.5\textwidth}
  \vspace{1.5em}
\begin{verbatim}
a, b |-
let w, x := mu_black(a) in
let y := mu_white(x, b) in
let z := eta_white() in
let := eta_black(y) in
z, w
\end{verbatim}
\end{minipage}

In the textual representation, I treat inputs a and b as free variables, while
outputs z and w are listed on the final line.
The rest of the expression is a series of \texttt{let}-expressions binding zero
or more variables to the results of an operator applied to other variables.
All of the variables are used linearly, avoiding any fan-out.

To turn this representation into a type system, I do the following.
First, I fix the usage annotation semiring as the
$\{\gr0, \gr1, \gr\omega\}$-semiring to achieve linearity.
Then, I set \AgdaBound{Ty} to be the following type.
All of the variables get type \AgdaInductiveConstructor{wire}, with a context
full of \AgdaInductiveConstructor{wire}-type variables corresponding to the
inputs of the graph.
Meanwhile, the whole expression has type
\AgdaInductiveConstructor{bundle}\AgdaSpace{}\AgdaBound{s}, where \AgdaBound{s}
gives the shape of the outputs.

\ExecuteMetaData[\Graphtex]{Sort}

The type system itself is as follows.
Each operator gets its own syntactic construct, with essentially the same naming
conventions as for the textual representation above.
Each of these constructs is an entire \texttt{let}-expression.
There is also an \AgdaInductiveConstructor{`end} construct, allowing us to
finish and list the outputs.

To see how the encodings of the operations work, let us look at the
\AgdaInductiveConstructor{`$\upmu\bullet$}-construct
($\tikz{\node (n) [circle,fill,minimum size=2mm,inner sep=0] {};%
  \draw[->] (n) to (2.12mm,-2.12mm); \draw[->] (n) to (-2.12mm,-2.12mm);%
  \draw[->] (0,3mm) to (n)}$).
It has two premises conjoined by $\sep$ (in fact, all premises in this language
are separated, with no use of sharing conjunction).
The first premise expects a simple value, corresponding to the input of the
operator.
The second premise is more complex, and represents the body/continuation of the
\texttt{let}-expression.
It binds two variables, corresponding to the two outputs, which can then be used
in later \texttt{let}-expressions or the outputs.
The bound variables all have type \AgdaInductiveConstructor{wire} and usage
annotation $\gr1$ (written \AgdaInductiveConstructor{u1} in Agda) to make them
behave linearly.

\ExecuteMetaData[\Graphtex]{qGraphL}
\ExecuteMetaData[\Graphtex]{GraphL}

The required premises of the \AgdaInductiveConstructor{`end}-rule are calculated
by \AgdaFunction{end-premises}.
This follows the tree structure of the shape, requiring a
\AgdaInductiveConstructor{wire}-term in an unextended context for each leaf.

\ExecuteMetaData[\Graphtex]{end-premises}

A graph is a term of the graph language.
Specifically, a graph with inputs of shape \AgdaBound{s} and outputs of shape
\AgdaBound{t} has the following type --- a term in a context of shape
\AgdaBound{s}, all of whose entries are wire variables with annotation $\gr1$,
whose result type is \AgdaInductiveConstructor{bundle}\AgdaSpace{}\AgdaBound{t}.

\ExecuteMetaData[\Graphtex]{Graph}

Finally, I give an example term \AgdaFunction{myGraph}.
As I said at the end of \cref{sec:terms}, writing a term out in full is tedious,
so I instead choose to use the machinery I discuss in detail in
\cref{sec:usage-elaborator}.
The \AgdaFunction{elab-unique} tool elaborates a well typed term into a well
typed and usage-correct term as long as it can infer assignments of usage
annotations that satisfy the constraints.
I add the prefix \texttt{u} to signify the unannotated (just well typed) terms.
All of the checks required by the elaboration procedure are done at Agda's type
checking time, so we know that the program listed below can be elaborated by
virtue of the whole Agda definition being type-checked.

The \AgdaFunction{uvar\#} tool is like the \AgdaFunction{var\#} tool I used
earlier --- allowing us to refer to variables by counting from the lefthand end
of the context.
I add these numbers to the picture of the graph for convenience.
Variables never go out of scope, despite being used, so these numbers do not
change with any bindings.

\noindent
\begin{minipage}[t]{0.33\textwidth}
  \begin{tikzpicture}[baseline]
    \node (i0) {};
    \node[right=2cm of i0] (i1) {};
    \node[below=1cm of i0] (mub) [circle,fill] {};
    \node[below right=1cm of mub] (muw) [circle,draw] {};
    \node[below left=1cm of muw] (etaw) [circle,draw] {};
    \node[below right=1cm of etaw] (etab) [circle,fill] {};
    \node[below left=1cm of etab] (o0) {};
    \node[below right=1cm of etab] (o1) {};
    \draw[->] (i0) to node[auto] {0} (mub);
    \draw[->] (i1) to[out=-90,in=45] node[auto] {1} (muw);
    \draw[->] (mub) to node[auto] {3} (muw);
    \draw[->] (muw) to node[above right] {4} (etab);
    \draw[->] (etaw) to node[auto] {5} (o0);
    \draw[->] (mub) to[out=-135,in=90] node[left,near start] {2} (o1);
  \end{tikzpicture}
\end{minipage}
\begin{minipage}[t]{0.67\textwidth}
  \AgdaNoSpaceAroundCode{}
  \ExecuteMetaData[\Graphtex]{myGraph}
  \AgdaSpaceAroundCode{}
\end{minipage}
